---
title: "PA1_template"
author: "Bikash Sarangi"
date: "3 August 2017"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Loading and preprocessing the data  


Code provided below helps read the csv file, eliminate scientific notation for numbers. The dataframe with the raw data read in is titled activity. 

### A. Load the data
```{r loader}
options(scipen = 999)
library(rmarkdown)
setwd("~/R/RepRes")
activity <-read.csv("activity.csv", stringsAsFactors = FALSE) 
```

### B. Process/transform the data (if necessary) into a format suitable for analysis

The 'date' field in the dataframe is transformed from a character to a date. This will help in the analysis later. A new dataframe called activityNew is created which exclues the rows with steps = NA. 

```{r process}
activity$date <- as.Date (as.POSIXct(activity$date, format = "%Y-%m-%d"))
activityNew <- na.omit(activity)
```

# 2. What is mean total number of steps taken per day?

For this part of the assignment, missing values in the dataset are ignored. So the dataframe activityNew is used.

### A. Make a histogram of the total number of steps taken each day

Histogram is created using gglpot2 package.

```{r plot}
library(ggplot2)
stepsM <- aggregate(activityNew$steps ~ with(activityNew, date), FUN = sum)
p <- ggplot(stepsM, aes(stepsM$`activityNew$steps`))+
  geom_histogram(fill =  "red", binwidth = 5000) +
  labs(x = "Steps ") + 
  labs(y= "Frequency") + 
  ggtitle("Total number of steps in a day")+
  theme(plot.title = element_text(hjust = 0.5))+
  scale_y_continuous(limits=c(0, 30))
suppressMessages(print(p))
```

### B. Calculate and report the mean and median total number of steps taken per day

```{r meanmed}
meanTime <- mean(stepsM$`activityNew$steps`)
medTime <- median(stepsM$`activityNew$steps`)
```
The mean number of steps is `r meanTime` while the median number of steps is `r medTime`.

# 3. What is the average daily activity pattern?

### A. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

Time series plot has been created using the base plotting system.

```{r interval}
intervalM <- aggregate(steps~interval, activityNew, FUN = mean)
plot (  
  intervalM$interval,
  intervalM$steps,
  type='l', 
  xlab= 'Interval', 
  ylab= 'Number of Steps',
  col='green',
  lwd=2)

```





### B. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

``` {r maxinterval}
pos <- which.max(intervalM$steps)
maxInt <- intervalM[pos ,1]
stepsInt <- intervalM[pos ,2]
``` 

The 5-minute interval, which on average across all the days in the dataset contains the maximum number of steps, is `r maxInt` with  `r stepsInt` steps.

# 4. Imputing missing values

### A. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missing}
missing <- nrow(activity) - nrow(activityNew)
```

The total number of missing values in the dataset (i.e. the total number of rows with NAs) is `r missing`.

### B. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. 

We have used the mean for that 5-minute interval for filling up missing data.

```{r missing2}
intervalMed <- aggregate(steps~interval, activityNew, FUN = mean)
size <- dim(activity)[1]
for (i in 1:size)
  {
  x <- activity[i,1]
  y[i] <- if (is.na(x)) {
                  subset(intervalMed, intervalMed$interval==activity[i,3])[1,2]
                } 
                else {
                  x}
}
```

### C. Create a new dataset that is equal to the original dataset but with the missing data filled in.


```{r newData}
activityFilled <- activity
activityFilled$stepsNew <- y
```

The new dataset is 'activityFilled'.

### D. Make a histogram of the total number of steps taken each day 

Histogram has been created using ggplot2 package.

```{r plotnew}
stepsMod <- aggregate(activityFilled$stepsNew ~ with(activityFilled, date), FUN = sum)
g <- ggplot(stepsMod, aes(stepsMod$`activityFilled$stepsNew`))+
     geom_histogram(fill =  "green", binwidth = 5000) +
     labs(x = "Steps ") + 
     labs(y= "Frequency") + 
     ggtitle("Total number of steps in a day - Missing Data Imputed")+
     theme(plot.title = element_text(hjust = 0.5))+
     scale_y_continuous(limits=c(0, 40))
suppressMessages(print(g))

```

### E. Calculate and report the mean and median total number of steps taken per day. 

```{r meanmedMod}
meanTimeMod <- mean(stepsMod$`activityFilled$stepsNew`)
medTimeMod <- median(stepsMod$`activityFilled$stepsNew`)
```
The mean number of steps is `r meanTimeMod` while the median number of steps is `r medTimeMod`.

### F. Do these values differ from the estimates from the first part of the assignment? 

Not materially.

### G. What is the impact of imputing missing data on the estimates of the total daily number of steps?

Mean remains the same while median has barely increased.

# Are there differences in activity patterns between weekdays and weekends?
```{r activity}
weekEnd <- function(d){
           z <- weekdays(as.Date(d))
           if (z %in% c("Saturday","Sunday")){
             "Weekend"
           }else
           {
            " Weekday" 
           }
} 
size <- dim(activityFilled)[1]
for (i in 1:size){
  activityFilled[i,5] <- weekEnd(activityFilled[i,2])
}

colnames(activityFilled)[5] <- "weekDay"
activityFilled$"weekDay" <- as.factor(activityFilled$"weekDay")
stepsFilled <- aggregate(stepsNew ~ weekDay+interval, activityFilled, FUN = mean)
gg <- ggplot(stepsFilled, aes(x = interval, y = stepsNew))+
     geom_line(color =  "green", size = 2) +
     labs(y= "Steps") + 
     ggtitle("Total number of steps in a day - Weekend Patterns with Missing Data")+
     scale_x_continuous("Interval", breaks = seq(min(stepsFilled$interval), max(stepsFilled$interval)+ 100, 200)) +
     theme(plot.title = element_text(hjust = 0.5))+
     facet_wrap(~ weekDay, scale ="fixed", nrow = 2)+
     theme_light()
suppressMessages(print(gg))



```